Cheminformatics, the Computer Science of Chemical Discovery, Turning Open-Source

### 2012-03-19 resubmit2/cheminformatics_second_CACM_revision.tex
final set of revisions

A fundamental principle of cheminformatics is that \emph{similar
molecules exhibit similar properties} \cite{Johnson:1990qf}. The
choice of representation is key in determining how such similarities
are evaluated, and thus the effectiveness of subsequent analyses. But
there is a further challenge in balancing computational costs with
the utility of a
representation. For example, a full 3D description of a molecule,
taking into account all possible conformers, would allow accurate
prediction of many properties. But the size of this representation and the
time required to evaluate it would be prohibitive. Instead, can we
obtain accurate predictions from a subset of conformers? Or, can we
obtain comparable accuracy by using a 2D representation? And if so,
what type of labels are required? Currently, many of these questions
are answered using trial and error through definition of an objective function
(usually root mean square error or percentage correct) and
iterative adaptation of descriptors and modeling approaches to
optimize the objective function. There is no
unifying theory to explain or suggest optimal approaches in all cases.

@@

Graph substructure matching is a variant of \emph{graph isomorphism},
which is widely believed to be computationally
intractable~\cite{cordella2001}. To execute a graph isomorphism search
across a full chemical database of thousands or even millions of
structures is unfeasible \cite{Weininger:2011ly}. Speedups can be
obtained via the use of heuristics such as structural
\emph{fingerprints}. Fingerprints encode characteristic features of a
given chemical structure, usually in a fixed-length
bitmap. Fingerprints fall broadly into two categories: structure keys
and hashed keys. In structure keys, each bit position corresponds to a
distinct substructure such as a functional group. Examples include MACCS and PubChem
keys. Hashed keys are those in which substructural patterns
are represented as strings and then hashed to a random bit
position. As a result, a given position can encode multiple
substructures. The advantage of such fingerprints is that they can
cover an arbitrarily large collection of substructures such as ``paths
of length $N$,'' or circular environments. Examples include Daylight fingerprints (which are ``folded'' to
optimize information density and screening speed) and
ECFPs (Extended Connectivity Fingerprints, which use local topological
information).
Given a binary fingerprint, we can first ``pre-screen'' a database, to
ignore molecules that cannot possibly match the query,
 by requiring that all bits in a query fingerprint must also
be present in the target fingerprint. Since the target fingerprints
are pre-computed, this check can be performed extremely
rapidly on modern hardware. As a result, we apply the actual
isomorphism test only on those molecules that pass the
screen. Fingerprints can also be used to rapidly search databases for
similar molecules, by using a similarity metric such as the
Tanimoto coefficient to compare the query and target
fingerprints. Additional heuristics have been developed
\cite{Swamidass:2007ve} to further speed up similarity searches.

@@

In many cases, the activity of a small molecule is due to its
interaction with a receptor. Traditionally, QSAR \cite{Hansch:1962vn,
 Free:1964ys} approaches do not take into account receptor features,
focusing only on small molecule features, and therefore lose valuable
information on ligand-receptor interactions. As a result, techniques
such as docking (which predicts how molecules fit together),
structure-based pharmacophore modeling (a 3D approach to capturing protein-ligand
interactions) and proteochemometric methods have been designed to take
into account both ligand and receptor structures. The last method is
an extension of statistical QSAR methods to simultaneously
model the receptor and ligand, as first reported by Lapinsh \textit{et
 al.}~\cite{lapinsh2001}.

The first step in the predictive modeling of biological activities is
to generate \emph{molecular descriptors} (a.k.a. \emph{features}) that
are numerical representations of different structural features. For
example, labeled graphs and their associated
characterizations are easily accessible to computer scientists, yet
such features miss significant physicochemical features (
properties such as charge, flexibility and so on). At the same time,
it can be difficult to objectively quantify many chemical aspects of a
molecule, such that the resultant descriptors are suitable for
predictive modeling.  Hence, the choice of a chemical descriptor
should by no means be treated as a ``solved'' problem. For a more
detailed discussion the reader is referred to more comprehensive
textbooks \cite{todeschini2000,faulon2010}.

As noted above, molecular graphs can be transformed to numerical
vector representations ranging from counts of elements to eigenvalues
of the Laplacian matrix.  Alternatively, we can compare molecular
graphs directly, via \emph{kernel methods}, where a kernel on graphs
$G$ and $G'$ provides a measure of how similar $G$ is to $G'$, or a
kernel on a single graph compares measured similarities between the
graph's nodes. In these cases, rather than compute vector
representations, we directly operate on the graph representations.
Both methods have advantages and disadvantages. The vector approach
requires one to identify a subset of \emph{relevant} (to the property
being modeled) descriptors: the feature selection problem is
well discussed in the data mining literature. A kernel approach
does not require feature selection, but one faces the problem of
evaluating a data set in a pairwise fashion, and must
identify an appropriate kernel. This is an important challenge as the
kernel should be selected to satisfy \emph{Mercer's condition} (a
well-known mathematical property in machine learning that makes a set
of observations easier to make predictions about), and this is not
always possible with traditional cheminformatics based kernels such as those based on multiple common substructures.
These challenges can
make kernel based methods prohibitive on larger datasets.

@@

One of the key challenges faced by predictive modeling is the fact
that small molecules are not static and do not exist in isolation.
Traditionally, predictive models have focused on a single structure
for a small molecule and ignored the receptor. Yet, small molecules
can exist in multiple tautomeric forms and conformations.
Enhancing the accuracy of predictions
will ideally require that the 3D geometries of that molecule be taken
into account and the receptor be considered as far as possible.
While some relevant low-energy conformers of small
molecules may be accessible in crystallographic databases, this is
not always the case. Though it is now possible to generate reasonable
low energy conformations \emph{ab initio}, the
``biologically relevant'' conformation might differ from the lowest
energy conformation of the molecule considered in isolation,
necessitating the need for conformational search.  Multi-conformer
modeling has been addressed in the 4D-QSAR methodology described by
Hopfinger \textit{et al.}  \cite{Albuquerque:1998ys}.  Recent
techniques such as \emph{multiple-instance learning} could also be
applied to the multi-conformer problem.

@@

Structure enumeration plays a fundamental role in \emph{molecular
  design} -- the design of compounds (drugs, for instance) that
optimize some physical, chemical, or biological property or activity
\cite{Schneider:2005uq}. A key challenge in this area is to combine
enumeration algorithms with efficient property prediction and is
closely related to methods in predictive modeling of chemical
properties (Section \ref{sec:activity-mining}). This problem is also
termed \emph{inverse QSAR}, where one must devise a set of features
that describe a molecule with a specified property, and subsequently
reconstruct the molecule from those features. As a side note, the
reconstruction of molecules from feature sets is can be considered a
cryptography problem. It is relevant to the pharmaceutical industry,
as it can be necessary to share data on molecules without sharing
explicit structures. Feature sets that allow easy reconstruction of
molecular structures are thus undesirable. A number of workers have
investigated the problem of identifying such one-way molecular
descriptors as well as methods to reconstruct molecules from
descriptors \cite{Masek:2008kx} and new methods that generate such
one-way molecular features but still correlate with physicochemical
properties would be valuable.

@@

Another relevant application of enumeration methods is the
\emph{generation of chemical reaction networks}. The problem here
consists of enumerating all the possible compounds that can be
produced by applying reaction rules to a set of initial molecules. By
reversing the reaction rules one can also find sets of starting
compounds necessary for producing a given target, known as
\emph{retrosynthesis}. Designing new drugs or chemicals, understanding
the kinetics of combustion and petroleum refining, studying the
dynamics of metabolic networks, applying metabolic engineering and
synthetic biology to produce heterologous compounds (compounds from
different species) in microorganisms, all involve enumeration of
reaction networks. As reviewed in Chapter 11 in \cite{faulon2010}
several network enumeration techniques have been developed. However,
these techniques generally suffer from a combinatorial explosion of
product compounds. One way to limit the number of compounds generated
is to simulate the dynamics of the network while it is being
constructed and remove compounds of low concentration. Following that
idea, methods have been developed based on the Gillespie Stochastic
Simulation Algorithm (SSA) to compute on-the-fly species
concentrations. Chemical reaction network enumeration and sampling is
an active field of research, particularly in the context of
metabolism, either to study biodegradation, or to propose metabolic
engineering strategies to biosynthesize compounds of commercial
interest. With metabolic network design, a difficulty is that in
addition to network generation based on reactions, one also needs to
verify that there are possible enzymatic events to enable reaction
catalysis (enzymes must be present to reduce the
energy required for some reactions to take place). That additional
task requires encompassing both chemical structures and protein
sequences and the development of tools that are at the
interface between cheminformatics and bioinformatics.

@@

Of equal importance to the development of new algorithms is freely
available data on which to train and test new methods. As has been
noted, recently large structure and bioactivity data collections have
become available. As a result, this enables much more robust
validation of cheminformatics methodologies as well as large scale
\emph{benchmarking of algorithms}. The latter is especially relevant
for the plethora of data mining techniques employed in
cheminformatics. At this point, the nature of Open Data focuses
primarily on structure and activity data types. On the other hand,
there is a distinct lack of textual data (such as journal articles)
that are ``Open''. While PubMed abstracts serve as a proxy
for journal articles, text mining methods in cheminformatics are
hindered by not being able to mine the full text of many scientific
publications.  Interestingly, patent information is publicly
accessible (cf. Google Patents) and represents an excellent resource
to support these types of efforts.

@@

There is an admitted learning curve, as significant domain
knowledge is required to attack many problems in cheminformatics due
to the underlying chemistry of the systems being studied. Indeed, many
issues faced by chemists do not admit as clean an abstraction as
``algorithms on a string,'' the way many bioinformatics algorithms can
be abstracted into a computer science framework.  Hence, while
one can certainly make significant contributions to cheminformatics by
just considering structures as graphs, one is limited to somewhat
abstract problems. Keeping in mind that cheminformatics is
fundamentally a practical field that serves to help and advance
experimental chemistry, the key challenges require an understanding of
the underlying chemical systems.

@@

\bibliographystyle{abbrv}
\bibliography{CACM_second_revision}

>>>>

A fundamental principle of cheminformatics is that \emph{similar
  molecules exhibit similar properties} \cite{Johnson:1990qf}. The
choice of representation is key in determining how such similarities
are evaluated, and thus the effectiveness of subsequent analyses. But
there is a further challenge in balancing computational costs with the
utility of a representation. For example, a full 3D description of a
molecule, taking into account all possible conformers, would allow
more accurate prediction of many properties (though this is also dependent
on qood quality force-fields, algorithms and so on). But the size of this
representation and the time required to evaluate it would be
prohibitive. Instead, can we obtain accurate predictions from a subset
of conformers? Or, can we obtain comparable accuracy by using a 2D
representation? And if so, what type of labels are required?
Currently, many of these questions are answered using trial and error
through definition of an objective function (usually root mean square
error or percentage correct) and iterative adaptation of descriptors
and modeling approaches to optimize the objective function. There is
no unifying theory to explain or suggest optimal approaches in all
cases.

@@

Graph substructure matching is a variant of \emph{graph isomorphism},
which is widely believed to be computationally
intractable~\cite{cordella2001}. To execute a graph isomorphism search
across a full chemical database of thousands or even millions of
structures is unfeasible \cite{Weininger:2011ly}. Speedups can be
obtained via the use of structural \emph{fingerprint}
filters. Fingerprints encode characteristic features of a given
chemical structure, usually in a fixed-length bitmap. Fingerprints
fall broadly into two categories: structure keys and hashed keys. In
structure keys, each bit position corresponds to a distinct
substructure such as a functional group. Examples include MACCS and
PubChem keys. Hashed keys are those in which substructural patterns
are represented as strings and then hashed to a random bit
position. As a result, a given position can encode multiple
substructures. The advantage of such fingerprints is that they can
cover an arbitrarily large collection of substructures such as ``paths
of length $N$,'' or circular environments. Examples include Daylight
fingerprints (which are ``folded'' to optimize information density and
screening speed) and ECFPs (Extended Connectivity Fingerprints, which
use local topological information).  Given a binary fingerprint, we
can first ``pre-screen'' a database, to ignore molecules that cannot
possibly match the query, by requiring that all bits in a query
fingerprint must also be present in the target fingerprint. Since the
target fingerprints are pre-computed, this check can be performed
extremely rapidly on modern hardware. As a result, we apply the actual
isomorphism test only on those molecules that pass the
screen. Fingerprints can also be used to rapidly search databases for
similar molecules, by using a similarity metric such as the Tanimoto
coefficient to compare the query and target fingerprints. Additional
heuristics have been developed \cite{Swamidass:2007ve} to further
speed up similarity searches.

@@

In many cases, the activity of a small molecule is due to its
interaction with a receptor. Traditionally, QSAR \cite{Hansch:1962vn,
  Free:1964ys} approaches do not consider receptor features, focusing
only on small molecule features, thus losing valuable information on
ligand-receptor interactions. As a result, techniques such as docking
(which predicts how molecules fit together), structure-based
pharmacophore modeling (a 3D approach to capturing protein-ligand
interactions) and proteochemometric methods have been designed to take
into account both ligand and receptor structures. The last method is
an extension of statistical QSAR methods to simultaneously model the
receptor and ligand, as first reported by Lapinsh \textit{et
  al.}~\cite{lapinsh2001}.

The first step in predicting biological activities is to generate
\emph{molecular descriptors} (a.k.a. \emph{features}) that are
numerical representations of structural features. For example, labeled
graphs and their associated characterizations are easily accessible to
computer scientists, yet such features miss significant
physicochemical features (properties such as surface distributions or
3D pharmacophores). At the same time, it can be difficult to
objectively quantify many chemical aspects of a molecule, such that
the resultant descriptors are suitable for predictive modeling.
Hence, the choice of a chemical descriptor should by no means be
treated as a ``solved'' problem. For a more detailed discussion the
reader is referred to more comprehensive textbooks
\cite{todeschini2000,faulon2010}.

Molecular graphs can be transformed to numerical vector
representations ranging from counts of elements to eigenvalues of the
Laplacian matrix.  Alternatively, molecular graphs can be compared 
directly via \emph{kernel methods}, where a kernel on graphs $G$ and
$G'$ provides a measure of how similar $G$ is to $G'$, or a kernel on
a single graph compares measured similarities between the graph's
nodes. In these cases, rather than compute vector representations, we
directly operate on the graph representations.  Both methods have
advantages and disadvantages. The vector approach requires one to
identify a subset of \emph{relevant} (to the property being modeled)
descriptors: the feature selection problem is well discussed in the
data mining literature. A kernel approach does not require feature
selection, but one faces the problem of evaluating a data set in a
pairwise fashion, and must identify an appropriate kernel. This is an
important challenge as the kernel should be selected to satisfy
\emph{Mercer's condition} (a well-known mathematical property in
machine learning that makes a set of observations easier to make
predictions about), and this is not always possible with traditional
cheminformatics based kernels such as those based on multiple common
substructures.  These challenges can make kernel based methods
prohibitive on larger datasets.

@@

One of the key challenges faced by predictive modeling is the fact
that small molecules are not static and do not exist in isolation.
Traditionally, predictive models have focused on a single structure
for a small molecule and ignored the receptor. Yet, small molecules
can exist in multiple tautomeric forms and conformations.  Enhancing
the accuracy of predictions will ideally require that the 3D
geometries of that molecule be taken into account and the receptor be
considered as far as possible.  Though it is now possible to generate
reasonable low energy conformations \emph{ab initio}, the
``biologically relevant'' conformation might differ significantly (in
terms of energetics) from the lowest energy conformation of the
molecule considered in isolation, necessitating the need for
conformational search.  Multi-conformer modeling has been addressed in
the 4D-QSAR methodology described by Hopfinger \textit{et al.}
\cite{Albuquerque:1998ys}.  Recent techniques such as
\emph{multiple-instance learning} could also be applied to the
multi-conformer problem.

@@

Structure enumeration plays a fundamental role in \emph{molecular
  design} -- the design of compounds that optimize some physical,
chemical, or biological property or activity
\cite{Schneider:2005uq}. A key challenge in this area is to combine
enumeration algorithms with efficient property prediction and is
closely related to methods in predictive modeling of chemical
properties (Section \ref{sec:activity-mining}). This problem is also
termed \emph{inverse QSAR}, where one must devise a set of features
that describe a molecule with a specified property, and subsequently
reconstruct the molecule from those features. As a side note, the
reconstruction of molecules from feature sets is can be considered a
cryptography problem. It is relevant to the pharmaceutical industry,
as it can be necessary to share data on molecules without sharing
explicit structures. Feature sets that allow easy reconstruction of
molecular structures are thus undesirable. A number of workers have
investigated the problem of identifying such one-way molecular
descriptors as well as methods to reconstruct molecules from
descriptors \cite{Masek:2008kx} and new methods that generate such
one-way molecular features but still correlate with physicochemical
properties would be valuable.

@@

Another relevant application of enumeration methods is the
\emph{generation of chemical reaction networks}. The problem here
consists of enumerating all the possible compounds that can be
produced by applying reaction rules to a set of initial molecules. By
reversing the reaction rules one can also find sets of starting
compounds necessary for producing a given target, known as
\emph{retrosynthesis}. Designing new drugs or chemicals, understanding
the kinetics of combustion and petroleum refining, studying the
dynamics of metabolic networks, applying metabolic engineering and
synthetic biology to produce heterologous compounds (compounds from
different species) in microorganisms, all involve enumeration of
reaction networks. As reviewed in Chapter 11 in \cite{faulon2010}
several network enumeration techniques have been developed but they
generally suffer from a combinatorial explosion of product
compounds. One way to limit the number of compounds generated is to
simulate the dynamics of the network while it is being constructed and
remove compounds of low concentration. Following that idea, methods
have been developed based on the Gillespie Stochastic Simulation
Algorithm (SSA) to compute on-the-fly species concentrations. Chemical
reaction network enumeration and sampling is an active field of
research, particularly in the context of metabolism, either to study
biodegradation, or to propose metabolic engineering strategies to
biosynthesize compounds of commercial interest. With metabolic network
design, a difficulty is that in addition to network generation based
on reactions, one also needs to verify that there are possible
enzymatic events to enable reaction catalysis (enzymes must be present
to reduce the energy required for some reactions to take place). That
additional task requires encompassing both chemical structures and
protein sequences and the development of tools that are at the
interface between cheminformatics and bioinformatics.

@@

Of equal importance to the development of new algorithms is freely
available data on which to train and test new methods. As has been
noted, recently large structure and bioactivity data collections have
become available. As a result, this enables much more robust
validation of cheminformatics methodologies as well as large scale
\emph{benchmarking of algorithms}. The latter is especially relevant
for the plethora of data mining techniques employed in
cheminformatics. Currently Open Data focuses primarily on structure
and activity data types and there is a distinct lack of textual data
(such as journal articles) that are ``Open''. While PubMed abstracts
serve as a proxy for journal articles, text mining methods in
cheminformatics are hindered by not being able to mine the full text
of many scientific publications.  Interestingly, patent information is
publicly accessible and can be used resource to support these types of
efforts. Of course, Open Data does not explicitly address data quality
issues nor the problems with integrating data sources. In many cases,
manual curation is the only option to maintain high quality databases.

@@

There is an admitted learning curve, as significant domain
knowledge is required to attack many problems in cheminformatics 
%%due to the underlying chemistry of the systems being studied.
Indeed, many issues faced by chemists do not admit as clean an
abstraction as ``algorithms on a string,'' the way many bioinformatics
algorithms can be abstracted into a computer science framework.
Hence, while one can certainly make significant contributions to
cheminformatics by just considering structures as graphs, one is
limited to somewhat abstract problems. Keeping in mind that
cheminformatics is fundamentally a practical field that serves to help
and advance experimental chemistry, the key challenges require an
understanding of the underlying chemical systems.

@@

\bibliographystyle{abbrv}
\bibliography{resubmission}
